Owlifier: Creating OWL-DL Ontologies from Simple Spreadsheet-Based Knowledge Descriptions\tnoteref{t1}

### 2008-12-03 docs/pubs/ISEI6/main.tex
section 2 draft done

\begin{abstract}
  Discovery and integration of data is important in many ecological
  studies, especially those that concern broad-scale ecological
  questions. Discovery and integration is often a difficult and
  time-consuming task for researchers, in part because informal,
  ambiguous, and sometimes inconsistent terms are used to describe
  data semantics.  Ontologies can help address this problem by
  providing a means to define ecological concepts to more consistently
  annotate, relate, and search for data sets.  However, unlike in
  molecular biology or biomedicine, few ontology development efforts
  exist within ecology. Ontology development often requires
  considerable expertise in ontology languages and development tools,
  which is often a barrier for ontology creation in ecology. In this
  paper we address this problem by providing an approach for ontology
  creation that allows ecologists to use common spreadsheet tools to
  describe different aspects of an ontology.  We present conventions
  for creating, relating, and constraining concepts through
  spreadsheets, and provide software tools for converting these
  ontologies into equivalent OWL-DL representations. We also consider
  inverse translations, i.e., to convert ontologies represented using
  OWL-DL into our spreadsheet format. Our approach allows large lists
  of terms to be easily related and organized into concept
  hierarchies, and generally provides a more intuitive and natural
  interface for ontology development by ecologists.
\end{abstract}

@@

Within the fields of molecular biology and biomedicine considerable
effort has gone into developing ontologies for improving data
discovery and integration
\cite{bard04:_ontol_in_biolog,ashburner00:_gene_ontol}. While similar
benefits can be obtained for ecological data, far fewer efforts exist
to develop richer and more consistent terminologies of ecology
concepts
\cite{madin08:_advan_ecolog_resear_with_ontol,parr20:_data_sharin_in_ecolog_and_evolut}.
The use of formal ontologies can significantly enhance metadata
descriptions of ecological data. Annotating data with ontology terms
can both help users interpret data as well as enable advanced
capabilities for data discovery and integration, e.g., by exploiting
subsumption and part-of hierarchies as well as more formal constraints
such as cardinality restrictions on properties and term equivalence.

Efforts to engage scientists in the development of ontologies
typically leverage the W3C Web Ontology Language (OWL)
\cite{smith04:_owl_web_ontol_languag_guide} as a standard XML syntax
for representing and sharing ontologies. A key advantage of OWL is
that it is supported by a wide range of generic tools, including
editors
\cite{knublauch04:_editin_descr_logic_ontol_with,kalyanpur05:_swoop},
reasoning systems
\cite{sirin07:_pellet,tsarkov06:_fact_descr_logic_reason}, query
languages
\cite{prudhommeaux08:_sparq_query_languag_for_rdf,motik05:_query_answer_for_owl_dl_with_rules},
and storage technologies
\cite{carroll04:_jena,broekstra02:_sesam}. Most of these tools,
however, are primarily targeted at experts in knowledge engineering
and software development familiar with the underlying description
logic semantics of OWL-DL \cite{grau08:_owl}. This is especially true
with ontology editors (such as Protege, SWOOP, etc.), which allow for
very detailed ontology specifications, but at the same time require
considerable understanding of the underlying ontology formalisms and
syntax. Thus, we see the lack of suitable ontology editing tools for
scientists as one of the major barriers for more wide-scale adoption
of ontologies in ecology.

@@

While not as expressive as OWL-DL, our approach can produce ontology
structures essential for enhanced data discovery and integration
\cite{madin07:_ontol_for_descr_and_synth}, while at the same time
provide a more accessible user interface for ecologists. Further, our
approach can be used to rapidly construct class hierarchies from long
lists of keywords using familiar spreadsheet software. For instance,
an ecologist can easily list (or import) a set of terms, and then
incrementally organize these into class hierarchies, and define
relevant properties and their constraints. Based on our experiences
using \owlifier\ with ecologists and evolutionary biologists studying
trait data, we found that this approach can enable them to quickly and
easily comprehend and construct useful ontologies.

@@

An \owlifier\ table defines an OWL-DL
\cite{smith04:_owl_web_ontol_languag_guide} ontology through a set of
\emph{blocks} representing one or more ontology definitions.  Each
non-empty row in an \owlifier\ table corresponds to a block. The type
of the block is given in the first column of the row. We describe each
type of block supported by \owlifier\ below. Here we assume that if
any properties or classes used in a block are not imported from
another ontology, then they are to be added to the ontology being
specified by the \owlifier\ table. In general, we name blocks
according to the more generic terms used in
\cite{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth}
instead of those of OWL-DL. This allows \owlifier\ to generate
ontologies that extend the observational model of
\cite{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth}
for data annotation, and in certain cases avoids confusion with
established terms commonly used within ecology.

@@

\myblock{Entity Blocks.} Entity blocks are the primary blocks used to
define ontologies. An entity block introduces new OWL classes and
specifies subclass relationships. Imported classes may also be used
within entity blocks by prefixing class names with namespace labels
(as described above).  Rows containing entity blocks take the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \textsf{entity} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 1$)
\end{itemize}
where each class $c_i$ is asserted in the current ontology to subsume
$c_{i+1}$, for $1 \le i < n$. That is, each $c_i$ in an entity block
induces the description logic axiom $c_{i+1} \sqsubseteq c_i$.  If
both $c_i$ and $c_{i+1}$ are imported classes, we say that the block
defines an ``articulation'' (i.e., mapping) between the two
classes. The following entity block defines a simple subclass
hierarchy.
\begin{tabbing}
  ~~\textsf{entity} \textsf{PhysicalFeature} 
    \textsf{AquaticPhysicalFeature} \textsf{River}
\end{tabbing}
This block states that Physical Feature, Aquatic Physical Feature, and
River are classes; River is a subclass of Aquatic Physical Feature;
and Aquatic Physical Feature is a subclass of Physical Feature. The
following entity block introduces a new class via an imported class.
\begin{tabbing}
  ~~\textsf{entity} \textsf{sweet:MarineEcosystem} 
    \textsf{IntertidalEcosystem}
\end{tabbing}
This block states that Intertidal Ecosystem is a subclass of the
Marine Ecosystem class imported from the SWEET ontology. Similarly,
assuming ``marine'' denotes an existing ontology of marine ecosystem
concepts, the following entity block defines a simple class
articulation.
\begin{tabbing}
  ~~\textsf{entity} \textsf{sweet:MarineEcosystem} 
    \textsf{marine:DeapSeaEcosystem}
\end{tabbing}
This block states that the Deap Sea Ecosystem class of the marine
ontology is a subclass of the Marine Ecosystem class of the SWEET
ontology.

\myblock{Synonym Blocks.} Synonym blocks define equivalence
relationships between ontology classes.  Rows containing synonym
blocks take the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \textsf{synonym} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where each class $c_i$ is equivalent to class $c_{i+1}$ in the current
ontology, for $1 \le i < n$. That is, each $c_i$ in a synonym blocks
induces a description logic axiom of the form $c_i \equiv
c_{i+1}$. The following synonym block defines a simple equivalence
relationship.
\begin{tabbing}
  ~~\textsf{synonym} \textsf{Maize} \textsf{Corn}
\end{tabbing}
This block states that the Maize and Corn classes are synonyms, i.e.,
equivalent classes. Similar to entity blocks, synonym blocks often
contain imported classes, e.g., to extend an existing ontology or to
define ontology articulations.

\myblock{Overlap Blocks.} Except in certain situations (described
further in \secref{sec:characteristics}), classes are generally
assumed to be disjoint in \owlifier.  Overlap blocks explicitly relax
this assumption by stating that a give set of classes may have
overlapping instances. Rows containing overlap blocks take the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \textsf{overlap} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where each class $c_i$ is allowed to share instances with each class
$c_j$, for $1 \le i,j \le n$. In particular, a given $c_i$ and $c_j$
in an overlap block are not defined to be disjoint classes in the
current ontology. As an example, consider the following entity blocks
that define the classes Estuary, Lagoon, and Marsh as subclasses of
Ecological Habitats.
\begin{tabbing}
  ~~\textsf{entity} \textsf{EcologicalHabitat} \textsf{Estuary} \\ 
  ~~\textsf{entity} \textsf{EcologicalHabitat} \textsf{Lagoon} \\ 
  ~~\textsf{entity} \textsf{EcologicalHabitat} \textsf{Marsh} 
\end{tabbing}
Given only these blocks, \owlifier\ treats Estuary, Lagoon, and Marsh
as disjoint classes. To relax this assumption and allow, e.g., types
of Lagoons to also be types of Estuaries, we explicitly add the
following overlap block
\begin{tabbing}
  ~~\textsf{overlap} \textsf{Estuary} \textsf{Lagoon}
\end{tabbing}
In general, overlap blocks are rarely used but provide a mechanism to
override the default behavior of \owlifier\ in asserting disjoint
classes.

@@

\myblock{Relationship Blocks.} Relationship blocks define
\emph{required} object properties of classes.  An object property
within OWL is a property defined between two class instances. Rows
containing relationship blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|}\hline \textsf{relationship} & $p$ & $c_1$
    & $c_2$ & \dots & $c_n$ \\ \hline
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where $p$ is an object property and each $c$ is a class.  For every
class $c_i$, the relationship block induces the DL axiom $c_i
\sqsubseteq \exists p . c_{i+1}$ stating that each instance of $c_i$
is $p$-related to some instance of $c_{i+1}$, for $1 \le i < n$.  For
example, the following block states that instances of the class
California Voles live in Grassy Areas.
\begin{tabbing}
  ~~\textsf{relationship} \textsf{livesIn} \textsf{CaliforniaVole} 
    \textsf{GrassyArea}
\end{tabbing}
In some cases, a particular property can apply to a sequence of
classes, and for convenience, each such class can be specified in a
single block. For example, consider the following block.
\begin{tabbing}
  ~~\textsf{relationship} \textsf{directlyBelow} \textsf{Hypolimnion} 
    \textsf{Thermocline} \textsf{Epilimnion} 
\end{tabbing}
This block states that, e.g., within a thermally stratified lake, the
Hypolimnion layer is directly below the Thermocline layer, and the
Thermocline layer is directly below the Epilimnion layer.

\myblock{Transitive Blocks.} Transitive blocks are special cases of
relationship blocks where the object property is asserted to be
transitive. If a property $p$ is declared to be transitive, whenever
$p$ relates an individual $o_1$ to an individual $o_2$, and an
individual $o_2$ to an individual $o_3$, then $p$ is also assumed to
relate $o_1$ to $o_3$. Rows containing transitive blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|}\hline \textsf{transitive} & $p$ & $c_1$
    & $c_2$ & \dots & $c_n$ \\ \hline
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where $p$ is an object property and each $c$ is a class.  The
following block is a simple example of a transitive relationship.
\begin{tabbing}
  ~~\textsf{transitive} \textsf{hasPart} \textsf{Body}
  \textsf{Head} \textsf{Eye} \textsf{Retina}
\end{tabbing} 
This block states that every instance of the class Body has a Head as
a part, every instance of the class Head has an Eye as a part, and
every instance of the class Eye has a Retina as a part. Moreover,
because the \textsf{hasPart} property above is defined to be
transitive, it is possible to infer that every instance of Body also
has an Eye and a Retina as a part through the implied relationship
restrictions $\textsf{Body} \sqsubseteq \exists \textsf{hasPart}
. \textsf{Head}$, $\textsf{Head} \sqsubseteq \exists \textsf{hasPart}
. \textsf{Eye}$, and $\textsf{Eye} \sqsubseteq \exists
\textsf{hasPart} . \textsf{Retina}$.

@@

An exact block ensures both a minimum and maximum number $m$ of
properties $p$ to distinct individuals that an individual of a class
must have. Rows containing exact blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{exact} & $p$ & $m$ & 
    $c_1$ & $c_2$ & $\dots$ & $c_n$
    \\ \hline
  \end{tabular} \hfill $(n \ge 2)$
\end{itemize}
where $m$ is the number of properties $p$ that instances of concept
$c_i$ must have to instances of concept $c_{i+1}$, for $1 \le i < n$.
For each class $c_i$, an exactly block induces the description logic
axiom $c_i \sqsubseteq \; (= m) \; p.c_{i+1}$ stating that each
instance of $c_i$ must be $p$-related to at $m$ unique instances of
$c_{i+1}$.

\myblock{Inverse Blocks.} Inverse blocks state that two object
properties are inverses of each other. If $p_1$ and $p_2$ are defined
to be inverse properties, whenever $p_1$ relates an individual $o_1$
to an individual $o_2$ then $p_2$ (as the inverse of $p_1$) is assumed
to relate $o_2$ to $o_1$.  Rows containing inverse blocks take the
form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|}\hline \textsf{inverse} & $p_1$ & $p_2$
\\ \hline
  \end{tabular}
\end{itemize}
where $p_1$ and $p_2$ are object properties. A common example of
inverse properties are \texttt{hasPart} and \texttt{partOf}, i.e., if
an individual $o_1$ has an individual $o_2$ as a part, then $o_2$ is
by definition a part of $o_1$.

@@

\myblock{Sufficient Blocks.} Sufficient blocks are similar to synonym
blocks in that they state equivalences between classes. However,
equivalence relationships are defined in sufficient blocks through
property restrictions (i.e., via anonymous classes). We consider two
main types of sufficient blocks. Rows containing sufficient blocks
take one of the following forms.
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|}\hline 
    \textsf{sufficient} & \textsf{min}, \textsf{max}, or 
    \textsf{exact} & $m$ & $c_1$ & $p$ & $c_2$\\ \hline
  \end{tabular}
\item[]
  \begin{tabular}{|l|l|l|l|l|}\hline 
    \textsf{sufficient} & \textsf{only} & $c_1$ & $p$ & \{\textsf{not}\} $c_2$\\ \hline
  \end{tabular}
\end{itemize}
The first form states that individuals that are $p$-related to at
least, at most, or exactly $m$ individuals of class $c_2$ are by
definition individuals of the class $c_1$. Sufficient blocks of this
form induce either the description logic axiom $c_1 \equiv \; (\le m)
\; p . c_2$, $c_1 \equiv \; (\ge m) \; p . c_2$, or $c_1 \equiv \; (=
m) \; p . c_2$ depending on whether the qualifier \textsf{min},
\textsf{max}, or \textsf{exact} is used, respectively. The second form
states that individuals that are $p$-related to only individuals of
class $c_2$ are by definition individuals of the class
$c_1$. Sufficient blocks of this form induce description logic axioms
$c_1 \equiv \forall p . c_2$. The complement of $c_2$ may also be used
within these blocks via the expression \textsf{not $c_2$}. Sufficient
blocks of this form induce description logic axioms $c_1 \equiv
\forall p . \neg c_2$ stating that individuals that are $p$-related to
only individual that are not of class $c_2$ (i.e., the complement of
$c_2$) are by definition individuals of class $c_1$.  Multiple
sufficient blocks for a class $c_1$ are treated as a single
description logic axiom, e.g., if each of the above blocks are defined
for $c_1$ we have $c_1 \equiv \; (\le m \; p .  c_2) \sqcap (\ge m \;
p . c_2) \sqcap (\forall p.c_2)$. The following blocks are simple
examples of sufficient conditions.
\begin{tabbing}
  ~~\textsf{sufficient} \textsf{mimimum} 1 \textsf{MammalProper} 
  \textsf{hasPart} \textsf{MammalProperEar}
  ~~\textsf{sufficient} \textsf{maximum} 2 \textsf{MammalProper} 
  \textsf{hasPart} \textsf{MammalProperEar}
  ~~\textsf{sufficient} \textsf{mimimum} 1 \textsf{MammalProper} 
  \textsf{hasPart} \textsf{Hair}

  ~~\textsf{sufficient} \textsf{mimimum} 3 \textsf{MammalProper} 
  \textsf{hasPart} \textsf{MiddleEarBone}
\end{tabbing}

\begin{tabbing}
  ~~\textsf{sufficient} \textsf{relationship} \textsf{MammalProper} \textsf{hasPart} \textsf{SweatGland} \\
  ~~\textsf{sufficient} \textsf{MammalProper} \textsf{hasPart} \textsf{Hair}
\end{tabbing}
These blocks state that

\myblock{Description Blocks.} Description blocks assign plain-text
definitions to concepts and properties. A description block has the
form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|}\hline \textsf{description} & $T$ & $S$
    \\ \hline
  \end{tabular}
\end{itemize}
where $T$ is either a property or a concept and $S$ is a description
string.

>>>>

\begin{abstract}
  Discovery and integration of data is important in many ecological
  studies, especially those that concern broad-scale ecological
  questions. Data discovery and integration is often a difficult and
  time-consuming task for researchers, which is due in part to the use
  of informal, ambiguous, and sometimes inconsistent terms for
  describing data content.  Ontologies offer a solution to this
  problem by providing consistent definitions of ecological concepts
  that in turn can be used to annotate, relate, and search for data
  sets.  However, unlike in molecular biology or biomedicine, few
  ontology development efforts exist within ecology. Ontology
  development often requires considerable expertise in ontology
  languages and development tools, which is often a barrier for
  ontology creation in ecology. In this paper we describe an approach
  for ontology creation that allows ecologists to use common
  spreadsheet tools to describe different aspects of an ontology.  We
  present conventions for creating, relating, and constraining
  concepts through spreadsheets, and provide software tools for
  converting these ontologies into equivalent OWL-DL
  representations. We also consider inverse translations, i.e., to
  convert ontologies represented using OWL-DL into our spreadsheet
  format. Our approach allows large lists of terms to be easily
  related and organized into concept hierarchies, and generally
  provides a more intuitive and natural interface for ontology
  development by ecologists.
\end{abstract}

@@

Within the fields of molecular biology and biomedicine considerable
effort has gone into developing ontologies for improving data
discovery and integration
\cite{ashburner00:_gene_ontol,bard04:_ontol_in_biolog}. While similar
benefits can be obtained for ecological data, far fewer efforts exist
to develop broad and consistent terminologies within ecology
\cite{madin08:_advan_ecolog_resear_with_ontol,parr20:_data_sharin_in_ecolog_and_evolut}.
The use of formal ontologies can significantly enhance metadata
descriptions of ecological data. For instance, annotating data with
ontology terms can both help users interpret data as well as enable
advanced capabilities for data discovery and integration, e.g., by
exploiting subsumption and part-of hierarchies as well as more formal
constraints such as cardinality restrictions on properties and term
equivalence \cite{madin08:_advan_ecolog_resear_with_ontol}.

Efforts to engage scientists in the development of ontologies
typically leverage the W3C Web Ontology Language (OWL)
\cite{smith04:_owl_web_ontol_languag_guide} as a standard XML syntax
for representing and sharing ontologies. A key advantage of OWL is
that it is supported by a wide range of generic tools, including
editors
\cite{knublauch04:_editin_descr_logic_ontol_with,kalyanpur05:_swoop},
reasoning systems
\cite{sirin07:_pellet,tsarkov06:_fact_descr_logic_reason}, query
languages
\cite{prudhommeaux08:_sparq_query_languag_for_rdf,motik05:_query_answer_for_owl_dl_with_rules},
and storage technologies
\cite{carroll04:_jena,broekstra02:_sesam}. Most of these tools,
however, are primarily targeted at experts in knowledge engineering
and software development familiar with the underlying description
logic semantics of OWL-DL \cite{grau08:_owl}. This is especially true
with ontology editors (such as Protege, SWOOP, etc.), which allow for
very detailed ontology specifications, but at the same time require
considerable understanding of the underlying ontology formalisms and
syntax. We see the lack of suitable ontology editing tools for
scientists without expertise in knowledge representation as one of the
major barriers for more wide-scale adoption of ontologies in ecology.

@@

While not as expressive as OWL-DL, our approach can produce ontology
structures that are essential for improved data discovery and
integration \cite{madin07:_ontol_for_descr_and_synth}. Just as
important, because spreadsheet tools are frequently used by ecologists
to store and analyze data, \owlifier\ can provide ecologists with a
familiar and accessible user interface for ontology creation. This
approach also leverages the ability of spreadsheet tools to organize
and manipulate tabular data, e.g., allowing users to rapidly construct
class hierarchies from long lists of keywords.  In this way, an
ecologist can easily construct (or import) a set of terms, and then
incrementally organize these into class hierarchies, properties, and
constraints. In initial experiments with ecologists and evolutionary
biologists studying trait data, we found that \owlifier\ enabled them
to quickly and easily comprehend and construct useful ontologies.

@@

As described above, an \owlifier\ table defines an OWL-DL
\cite{smith04:_owl_web_ontol_languag_guide} ontology through a set of
\emph{blocks} representing one or more ontology definitions.  Each
non-empty row in an \owlifier\ table corresponds to a block. The type
of the block is given in the first column of the row. 
% This section describes in detail each type of block supported by
% \owlifier.
We assume that if any properties or classes used in a block are not
imported from another ontology, then they are to be added to the
ontology being specified by the \owlifier\ table (i.e., the
``current'' ontology). In general, we name blocks according to the
terms used in
\cite{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth}
as opposed to the names used for corresponding constructs in
OWL-DL. This choice of block names helps to simplify terminology
(e.g., we use ``relation'' below instead of ``object property''),
allows \owlifier\ to easilty generate ontologies that extend the
observational model of
\cite{bowers08:_concep_model_framew_for_expres,madin07:_ontol_for_descr_and_synth},
and avoids confusion with established terms commonly used within
ecology (e.g., ``class'').

@@

\myblock{Entity Blocks.} Entity blocks are the primary blocks used to
define ontologies. An entity block introduces new OWL classes and
specifies subclass relationships. Imported classes may also be used
within entity blocks by prefixing class names with namespace labels
(as described above).  Rows containing entity blocks take the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \textsf{entity} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 1$)
\end{itemize}
where each class $c_i$ is asserted in the current ontology to subsume
$c_{i+1}$, for $1 \le i < n$. That is, each $c_i$ in an entity block
induces the description logic axiom $c_{i+1} \sqsubseteq c_i$.  If
both $c_i$ and $c_{i+1}$ are imported classes, we say that the block
defines an ``articulation'' (i.e., mapping) between the two
classes. The following entity block defines a simple subclass
hierarchy.
\begin{tabbing}
  ~~\textsf{entity} \textsf{PhysicalFeature} 
    \textsf{AquaticPhysicalFeature} \textsf{River}
\end{tabbing}
This block states that Physical Feature, Aquatic Physical Feature, and
River are classes; River is a subclass of Aquatic Physical Feature;
and Aquatic Physical Feature is a subclass of Physical Feature. The
following entity block introduces a new class via an imported class.
\begin{tabbing}
  ~~\textsf{entity} \textsf{sweet:MarineEcosystem} 
    \textsf{IntertidalEcosystem}
\end{tabbing}
This block states that Intertidal Ecosystem is a subclass of the
Marine Ecosystem class imported from the SWEET ontology. Similarly,
assuming ``marine'' denotes an existing ontology of marine ecosystem
concepts, the following block defines a simple class articulation.
\begin{tabbing}
  ~~\textsf{entity} \textsf{sweet:MarineEcosystem} 
    \textsf{marine:DeapSeaEcosystem}
\end{tabbing}
This block states that the Deap Sea Ecosystem class of the marine
ontology is a subclass of the Marine Ecosystem class of the SWEET
ontology (thus defining a mapping between these two ontologies).

\myblock{Synonym Blocks.} Synonym blocks define equivalence
relationships between ontology classes.  Rows containing synonym
blocks take the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \textsf{synonym} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where each class $c_i$ is equivalent to class $c_{i+1}$ in the current
ontology, for $1 \le i < n$. That is, each $c_i$ in a synonym block
induces a description logic axiom of the form $c_i \equiv
c_{i+1}$. The following synonym block creates a simple equivalence
relationship.
\begin{tabbing}
  ~~\textsf{synonym} \textsf{Maize} \textsf{Corn}
\end{tabbing}
This block states that the Maize and Corn classes are synonyms
(equivalent classes). Similar to entity blocks, synonym blocks often
contain imported classes for extending existing ontologies or defining
ontology mappings.

\myblock{Overlap Blocks.} Except in certain situations (described
further in \secref{sec:characteristics}), classes are generally
assumed to be disjoint in \owlifier.  Overlap blocks explicitly relax
this assumption by stating that a given set of classes may have
overlapping instances. Rows containing overlap blocks take the form
\begin{itemize}
\item[] 
  \begin{tabular}{|l|l|l|l|l|}\hline
    \textsf{overlap} & $c_1$ & $c_2$ & \dots & $c_n$ \\ \hline 
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where each class $c_i$ is allowed to share instances with each class
$c_j$, for $1 \le i,j \le n$. That is, a given $c_i$ and $c_j$ in an
overlap block are not defined to be disjoint classes in the current
ontology. As an example, consider the following entity blocks that
define the classes Estuary, Lagoon, and Marsh as subclasses of
Ecological Habitats.
\begin{tabbing}
  ~~\textsf{entity} \textsf{EcologicalHabitat} \textsf{Estuary} \\ 
  ~~\textsf{entity} \textsf{EcologicalHabitat} \textsf{Lagoon} \\ 
  ~~\textsf{entity} \textsf{EcologicalHabitat} \textsf{Marsh} 
\end{tabbing}
Given only these blocks, \owlifier\ treats Estuary, Lagoon, and Marsh
as disjoint classes. To relax this assumption and allow, e.g., types
of Lagoons to also be types of Estuaries, we explicitly add the
following overlap block
\begin{tabbing}
  ~~\textsf{overlap} \textsf{Estuary} \textsf{Lagoon}
\end{tabbing}
In general, overlap blocks are rarely used but provide a mechanism to
override the default behavior of \owlifier\ in asserting disjoint
classes.

@@

\myblock{Relationship Blocks.} Relationship blocks define
\emph{required} class object properties.  An object property within
OWL is a property defined between two class instances. Rows containing
relationship blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|}\hline \textsf{relationship} & $p$ & $c_1$
    & $c_2$ & \dots & $c_n$ \\ \hline
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where $p$ is an object property and each $c$ is a class.  For every
class $c_i$, the relationship block induces the description logic axiom $c_i
\sqsubseteq \exists p . c_{i+1}$ stating that each instance of $c_i$
is $p$-related to some instance of $c_{i+1}$, for $1 \le i < n$.  For
example, the following block states that instances of the class
California Voles live in Grassy Areas.
\begin{tabbing}
  ~~\textsf{relationship} \textsf{livesIn} \textsf{CaliforniaVole} 
    \textsf{GrassyArea}
\end{tabbing}
In some cases, a particular property can apply to a sequence of
classes, and for convenience, each such class can be specified in
\owlifier\ using a single block. For example, consider the following
block.
\begin{tabbing}
  ~~\textsf{relationship} \textsf{directlyBelow} \textsf{Hypolimnion} 
    \textsf{Thermocline} \textsf{Epilimnion} 
\end{tabbing}
This block states that, e.g., within a thermally stratified lake, the
Hypolimnion layer is directly below the Thermocline layer, and the
Thermocline layer is directly below the Epilimnion layer.

\myblock{Transitive Blocks.} Transitive blocks are special cases of
relationship blocks where the object property is asserted to be
transitive. If a property $p$ is declared to be transitive, whenever
$p$ relates an individual $x$ to an individual $y$, and an
individual $y$ to an individual $z$, then $p$ is also assumed to
relate $x$ to $z$. Rows containing transitive blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|}\hline \textsf{transitive} & $p$ & $c_1$
    & $c_2$ & \dots & $c_n$ \\ \hline
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where $p$ is an object property and each $c$ is a class.  The
following block is a simple example of a transitive relationship.
\begin{tabbing}
  ~~\textsf{transitive} \textsf{hasPart} \textsf{Body}
  \textsf{Head} \textsf{Eye} \textsf{Retina}
\end{tabbing} 
This block states that every instance of the class Body has a Head as
a part, every instance of the class Head has an Eye as a part, and
every instance of the class Eye has a Retina as a part. Moreover,
because the \textsf{hasPart} property above is defined to be
transitive, it is possible to infer that every instance of Body also
has an Eye and a Retina as a part through the inherited relationship
restrictions $\textsf{Body} \sqsubseteq \exists \textsf{hasPart}
. \textsf{Head}$, $\textsf{Head} \sqsubseteq \exists \textsf{hasPart}
. \textsf{Eye}$, and $\textsf{Eye} \sqsubseteq \exists
\textsf{hasPart} . \textsf{Retina}$.

@@

An exact block ensures both a minimum and maximum number $m$ of
properties $p$ to distinct individuals that an individual of a class
must have. Rows containing exact blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|l|}\hline \textsf{exact} & $p$ & $m$ & 
    $c_1$ & $c_2$ & $\dots$ & $c_n$
    \\ \hline
  \end{tabular} \hfill $(n \ge 2)$
\end{itemize}
where $m$ is the number of properties $p$ that instances of concept
$c_i$ must have to instances of concept $c_{i+1}$, for $1 \le i < n$.
For each class $c_i$, an exact block induces the description logic
axiom $c_i \sqsubseteq \; (= m) \; p.c_{i+1}$ stating that each
instance of $c_i$ must be $p$-related to $m$ unique instances of
$c_{i+1}$.

\myblock{Inverse Blocks.} Inverse blocks state that two object
properties are inverses of each other. If $p_1$ and $p_2$ are defined
to be inverse properties, whenever $p_1$ relates an individual $x$
to an individual $y$ then $p_2$ (as the inverse of $p_1$) is assumed
to relate $y$ to $x$.  Rows containing inverse blocks take the
form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|}\hline \textsf{inverse} & $p_1$ & $p_2$
\\ \hline
  \end{tabular}
\end{itemize}
where $p_1$ and $p_2$ are object properties. A common example of
inverse properties are \textsf{hasPart} and \textsf{partOf}, i.e., if
an individual $x$ has an individual $y$ as a part, then $y$ is
by definition a part of $x$.

@@

\myblock{Sufficient Blocks.} Sufficient blocks are similar to synonym
blocks in that they state equivalences between classes. However,
equivalence relationships are defined in sufficient blocks through
property restrictions (i.e., via anonymous classes). We consider two
main types of sufficient blocks. Rows containing sufficient blocks
take one of the following forms.
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|l|l|}\hline 
    \textsf{sufficient} & \textsf{min}, \textsf{max}, or 
    \textsf{exact} & $m$ & $p$ & $c_1$ & $c_2$ & \dots & $c_n$\\ \hline
  \end{tabular} \hfill $(n \ge 2)$
\item[]
  \begin{tabular}{|l|l|l|l|l|l|l|}\hline 
    \textsf{sufficient} & \textsf{only} & $p$ & $c_1$ & $c_2$ & \dots & 
    \{\textsf{not}\} $c_n$\\ \hline
  \end{tabular} \hfill $(n \ge 2)$
\end{itemize}
The first form states that individuals $p$-related to at least, at
most, or exactly $m$ individuals of class $c_{i+1}$ are by definition
individuals of the class $c_i$, for $1 \le i < n$. Each class $c_i$ of
a sufficient block of this form induces either the description logic
axiom $c_i \equiv \; (\le m) \; p . c_{i+1}$, $c_i \equiv \; (\ge m)
\; p . c_{i+1}$, or $c_i \equiv \; (= m) \; p . c_{i+1}$ depending on
whether the qualifier \textsf{min}, \textsf{max}, or \textsf{exact} is
used, respectively.

The second form states that individuals $p$-related to only
individuals of class $c_{i+1}$ are by definition individuals of the
class $c_i$, for $1 \le i < n$. For each $c_i$ a sufficient block of
this form induces the description logic axiom $c_i \equiv \forall p
. c_{i+1}$. The complement of the last class in a sequence may also be
used within these blocks via the expression \textsf{not
  $c_n$}. Sufficient blocks of this form induce description logic
axioms $c_{n-1} \equiv \forall p . \neg c_n$ stating that individuals
that are $p$-related to only individuals that are not of class $c_n$
(i.e., the complement of $c_n$) are by definition individuals of class
$c_{n-1}$.  Multiple sufficient blocks for a class are treated as a
single description logic axiom through intersection. For example, the
blocks
\begin{tabbing}
  ~~\textsf{sufficient} \= \textsf{only} \= $m$ \= $c_1$ \= $c_2$ \= \kill
  ~~\textsf{sufficient} \> \textsf{min} \> $m$ \> $p$ \> $c_1$ \> $c_2$ \\ 
  ~~\textsf{sufficient} \> \textsf{max} \> $m$ \> $p$ \> $c_1$ \> $c_2$ \\ 
  ~~\textsf{sufficient} \> \textsf{only} \> $p$ \> $c_1$ \> $c_2$
\end{tabbing}
result in the single axiom $c_1 \equiv \; (\le m \; p .  c_2) \sqcap
(\ge m \; p . c_2) \sqcap (\forall p.c_2)$. The following blocks are
simple examples of sufficient conditions.
\begin{tabbing}
  ~~\textsf{sufficient} \textsf{exact} \= \kill
  ~~\textsf{sufficient} \textsf{min} \> 1 \textsf{hasPart} 
  \textsf{Mammal} \textsf{Hair} \\
  ~~\textsf{sufficient} \textsf{exact} \> 2  \textsf{hasPart} 
  \textsf{Mammal} \textsf{MammalianEar} \\
  ~~\textsf{sufficient} \textsf{only} \> \textsf{feedsYoungWith} 
  \textsf{Mammal} \textsf{MammaryGland}\\
  ~~\textsf{sufficient} \textsf{exact} \> 3 \textsf{hasPart} 
  \textsf{MammalianEar} \textsf{MiddleEarBone}
\end{tabbing}
The first three blocks state that individuals that have Hair as a
part, that have exactly two Mammalian Ears as parts, and that feed
their young via Mammary Glands are Mammals. The last block states
individuals that have exactly three Middle Ear Bones are Mammalian
Ears.
